D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2024-01-06 ootavacuous.tex
OOTA definitions and I/O: Complete the thought
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Wordsmith "what about non-volatile" section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

In addition, I/O operations are observed behavior, which greatly restricts
optimizations involving them.

@@

@@

\emph{Because of unsolved mathematical problems, precise definitions of
``semantic dependency'' are problematic.
Therefore, C++ implementations must assume that a given dependency is
a semantic dependency unless and until it can prove otherwise.}

@@

>>>>

In addition, I/O operations are observed behavior, which greatly restricts
optimizations involving them, in turn reducing the need to evaluate
potential semantic dependencies.

@@

Many analysis tools would respond to these examples by failing, for example,
by consuming all available CPU and memory.

@@

\emph{Because of unsolved mathematical problems, precise definitions of
``semantic dependency'' have limitations.
Therefore, C++ implementations must assume that a given dependency is
a semantic dependency unless and until it can prove otherwise.}

@@

\subsection{But What About Non-Volatile Atomics?}
\label{sec:But What About Non-Volatile Atomics?}

Volatile atomic accesses have some very attractive properties from
the viewpoint of analyzing executions for OOTA behavior because C++
compilers are prohibited from reordering, omitting, fusing, or inventing
volatile accesses.
Thus, if the compiler emits code that faithfully reflects the behaviors
specified by the source code, the hardware memory model will prohibit
OOTA cycles.

However, there is some question as to what exactly a C++ implementation
is permitted to do with non-volatile atomic accesses~\cite{JFBastien2015N4455}.
For example,
Appendix~\ref{app:Inventing Atomic Loads}
shows a clever litmus test that results in OOTA-like behavior
if a C++ implementation duplicates an atomic load, compares the two values
loaded, and then, if the two loads return different values, chooses the
one that eliminates a semantic dependency that would otherwise remain
in place.
Perhaps in the fullness of time, such examples will be considered to be
simple optimizations rather than full-on OOTA cycles, but in the meantime
C++ implementations might wish to prohibit such behavior.

One way to accomplish this while still permitting optimization of code
containing non-volatile atomics is to model non-volatile behavior on that
of volatile atomics, but with these specific weakenings:

\begin{itemize}
\item	Loads that do not affect observable behavior can be omitted,
	along with any computations based on their return values.
\item	Given a pair of relaxed loads from the same object that can
	be reordered to be adjacent to each other, one of the pair can
	be omitted with uses of the omitted load instead using the value
	returned by the surviving load.
	This is termed \emph{load fusing}.
\item	Given a pair of relaxed stores to the same object that can
	be reordered to be adjacent to each other, the first of the pair
	can be omitted.
	This is termed \emph{store fusing}.
\item	If an expression can be proven to always result in a constant,
	that constant can be substituted for that expression.
	Note well that combining this with specialization optimizations
	can convert address and data dependencies to weaker control
	dependencies.
	Address and data dependencies order later loads and stores,
	and control dependencies order only later stores.\footnote{
		Yes, yes, the C++ standard does not mention address or
		data dependencies, and mandates control dependencies
		only indirectly.
		Such dependencies nevertheless still exist.}
\end{itemize}

Note that load and store fusing can only be applied a finite number of
times.
In practice, this means that fusing can be used to unroll loops, but
not to hoist atomic loads or stores out of loops than cannot be proven
to be always finite.

These optimizations are examples, not necessarily a full set.
Additional optimizations might be added over time, but great care will
of course be required.
